---
title: "Dancing Landscape Demo"
author: "Joseph Abolt"
date: "April 6, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Libraries
```{r}
# install.packages('knitr', repos = c('http://rforge.net', 'http://cran.rstudio.org'), type = 'source')
# install.packages("ggplot2")
# install.packages("gganimate")
# install.packages("gifski")
# install.packages("png")
# install.packages("transformr")
library("knitr")
library("ggplot2")
library("gganimate")
library("gifski")
library("png")
library("transformr")
```

## Dancing Landscape

* Landscape: term for any multidimensional surface used for optimization
* Dancing Landscape: a landscape that moves over time

### Sources of Dancing Landscapes

* Complex behavior
    * When factors that create landscapes are interdependent
    * When factors that create landscapes change over time

### Examples of Dancing Landscapes

* The stock market
* Customer demand by product tag over time

## Demonstration
``` {r}
f2D <- function(x, t) {
  return (cos(x) * sin(t * pi) - (abs(x)/10) + x/20)
}

par(mfrow=c(3, 3))
plot(f2D(seq(-10, 10, 0.1), 0.00), type="l")
plot(f2D(seq(-10, 10, 0.1), 0.17), type="l")
plot(f2D(seq(-10, 10, 0.1), 0.33), type="l")
plot(f2D(seq(-10, 10, 0.1), 0.50), type="l")
plot(f2D(seq(-10, 10, 0.1), 0.67), type="l")
plot(f2D(seq(-10, 10, 0.1), 0.83), type="l")
plot(f2D(seq(-10, 10, 0.1), 1.00), type="l")
plot(f2D(seq(-10, 10, 0.1), 1.17), type="l")
plot(f2D(seq(-10, 10, 0.1), 1.33), type="l")
par(mfrow=c(1, 1))
```

## Assemble Sample dataframe
```{r}
constructData2D <- function (func, xMin, xMax, precision, timeMin, timeMax)
{
  # Interval is bounded by gganimate - this just works, do not ask questions
  timeInterval = ((timeMax - timeMin) / 50) + 0.001
  
  data <- data.frame(X=0, Y=0, Time=0)
  i = 1
  for (t in seq(timeMin, timeMax, timeInterval))
  {
    for (x in seq(xMin, xMax, precision))
    {
      data[i,]$X = x
      data[i,]$Y = func(x, t)
      data[i,]$Time = t
      i = i + 1
    }
  }
  
  return (data)
}

timeData = constructData2D(f2D, -10, 10, 0.2, 0, 5)

## Output the dataframe and the graph it represents
str(timeData)
ggplot(timeData, aes(X, Y, group=Time)) +
  geom_path()
```

## Animation of function over time
```{r}
visualizeFunction <- function(dataframe)
{
  animation = ggplot(dataframe, aes(X, Y, group=Time)) +
    geom_path() +
    transition_states(Time, transition_length=1, state_length=1, wrap=F) +
    labs(title = "Time: {closest_state}") +
    enter_fade() + 
    exit_fade()
  
  if (!is.null(dataframe$OptimalX))
  {
    animation = animation +
      geom_point(x=dataframe$OptimalX, y=dataframe$OptimalY, colour="hotpink", size=4)
  }
  
  animation
}

visualizeFunction(timeData)
```

## Setup Hill Climbing
The function updates the OptimalX and OptimalY columns of a dataframe containing X, Y, and Time to include the best results from hill climbing optimization function. It attempts to find the maxima.

ticksPerUnitTime controls how fast the hill climbing runs compared to the function. For every increase in the time value of 1.0, the hill climbing will update its value ticksPerUnitTime times.

costFunction is the function being optimized, of the form y = costFunction(x, time).

At time = 0, the algorithm will use a single random guess. That is, do not expect the first value to be a good one. Setting startX can control this instead of a random value.

windowSize controls how far the hill climb algorithm can guess for each guess. It will look up to (windowSize / 2) left or right from its current best
```{r}
hillClimb <- function(timeDataFrame, ticksPerUnitTime, costFunction, startX = NULL, windowSize = 1)
{
  # Setup variables
  currentTime = min(timeDataFrame$Time)
  currentTick = 1
  currentSubset = timeDataFrame[timeDataFrame$Time == currentTime,]
  bestX = startX
  if (is.null(bestX))
  {
    bestX = runif(1, min(currentSubset$X), max(currentSubset$X))
  }
  bestY = costFunction(bestX, currentTime)
  width = windowSize / 2
  getOption <- function() { 
    return (runif(1, 
              max(bestX - width, min(currentSubset$X)), 
              min(bestX + width, max(currentSubset$X))))
  }

  # Record optimal x and y
  timeDataFrame[timeDataFrame$Time == currentTime,]$OptimalX = rep(bestX, nrow(currentSubset))
  timeDataFrame[timeDataFrame$Time == currentTime,]$OptimalY = rep(bestY, nrow(currentSubset))

  # Iterate through time
  nextSubset = timeDataFrame[timeDataFrame$Time > currentTime,]
  while (nrow(nextSubset) > 0)
  {
    # Calculate number of hill climb ticks (guesses) to do
    nextTime = min(nextSubset$Time)
    iterations = floor(nextTime * ticksPerUnitTime) - currentTick
    
    # Update variables
    currentTime = nextTime
    currentSubset = timeDataFrame[timeDataFrame$Time == currentTime,]
    bestY = costFunction(bestX, currentTime)

    # Actually do hill climbing
    if (iterations >= 1)
    for (i in 1:iterations)
    {
      currentTick = currentTick + 1
      testX = getOption()
      testY = costFunction(testX, currentTime)
      if (testY > bestY)
      {
        bestX = testX
        bestY = testY
      }
    }
    
    # Record new optimal x and y
    timeDataFrame$OptimalX[timeDataFrame$Time == currentTime] = rep(bestX, nrow(currentSubset))
    timeDataFrame$OptimalY[timeDataFrame$Time == currentTime] = rep(bestY, nrow(currentSubset))
    
    # Load next time subset
    nextSubset = timeDataFrame[timeDataFrame$Time > currentTime,]
  }
  
  return (timeDataFrame)
}
```

## Run and Visualize Hill Climbing
```{r}
# Approximately 1 guess per tick
timeData = hillClimb(timeData, 9, f2D, startX = -10)
visualizeFunction(timeData)
```

```{r}
# Approximately 0.3 guesses per tick
timeData = hillClimb(timeData, 3, f2D, startX = -10)
visualizeFunction(timeData)
```

```{r}
# Approximately 3 guesses per tick
timeData = hillClimb(timeData, 28, f2D, startX = -10)
visualizeFunction(timeData)
```

## Takeaways
 * Optimization is "easy" when optimzer speed > function change
 * Optimization is difficult when optimizer speed < function change
 