---
title: "Workshop"
author: "Joseph Abolt"
date: "April 21, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# install.packages('knitr', repos = c('http://rforge.net', 'http://cran.rstudio.org'), type = 'source')
# install.packages("ggplot2")
# install.packages("gganimate")
# install.packages("gifski")
# install.packages("png")
# install.packages("transformr")
library("knitr")
library("ggplot2")
library("gganimate")
library("gifski")
library("png")
library("transformr")
```

# Complexity Lecture Workshop
This sets up an agent-based model that exhibits complexity. We will use it to for the following workshop, where you will attempt to optimize over the output from the model.

The model is a square grid of cells, wrapping on both axes, with one agent per cell. Each agent has a value that it can adjust up or down. An agent wants to have a higher value than three of its neighbors, but does not want to be the highest. Agents will:

 * Subtract 3 if they are the highest
 * Add 1 if they are below average
 * Add 1 if they are above average and not the second-highest

### Setup Utility Functions
```{r}
# From https://stackoverflow.com/questions/2453326/fastest-way-to-find-second-third-highest-lowest-value-in-vector-or-column
maxN <- function(x, N=2){
  len <- length(x)
  if(N>len){
    warning('N greater than length(x).  Setting N=length(x)')
    N <- length(x)
  }
  sort(x,partial=len-N+1)[len-N+1]
}
# End sourced code

# Gets the change the agent performs based on its neighbors (not the new value)
getAgentChange <- function(agentValue, neighborValues)
{
  neighborValues = c(neighborValues, agentValue)
  if (agentValue >= max(neighborValues))
    return (-3)
  if (agentValue < mean(neighborValues))
    return (1)
  if (agentValue >= mean(neighborValues) && agentValue < maxN(neighborValues, 2))
    return (1)
  return (0)
}
```

### Setup grid
```{r}
# Builds a world with the specified height and width.
# startingValues should be a vector, with one value 
# per cell (applied down each column, then across each row).
buildWorld <- function(height, width, startingValues = NULL)
{
  if (is.null(startingValues))
  {
    startingValues = rep(100, height * width)
  }
  densityValue = 1 / (height * width)
  sumStartValues = sum(startingValues)
  world <- data.frame(X=0, Y=0, Value=startingValues[1], Density = densityValue)
  i = 1
  for (x in 1:width)
  {
    for (y in 1:height)
    {
      world[i,]$X = x
      world[i,]$Y = y
      world[i,]$Value = startingValues[i]
      world[i,]$Density = startingValues[i] / sumStartValues
      i = i + 1
    }
  }
  
  return (world)
}

# Create and graph an initial world
world = buildWorld(16, 16, seq(100, 125.6, 0.1))

ggplot(world, aes(X, Y, z = Density)) +
  geom_raster(aes(fill = Density)) +
  geom_contour(colour = "white")
```

### Setup Changing Values
```{r}
# Updates the passed in world according to the updateFunction
# Neighbors are only the adjacent four. If includeDiagonals is
# true, then all eight adjacent cells are counted as neighbors.
updateWorld <- function(world, updateFunction, includeDiagonals = FALSE)
{
  newWorld = buildWorld(max(world$X), max(world$Y))
  for (x in 1:max(world$X))
  {
    xLower = x - 1
    if (xLower == 0) xLower = max(world$X)
    xUpper = x + 1
    if (xUpper > max(world$X)) xUpper = 1
    
    for (y in 1:max(world$Y))
    {
      yLower = y - 1
      if (yLower == 0) yLower = max(world$Y)
      yUpper = y + 1
      if (yUpper > max(world$Y)) yUpper = 1
      
      neighborValues = c()
      neighborValues = c(neighborValues, world[world$X == xUpper & world$Y == y,]$Value)
      neighborValues = c(neighborValues, world[world$X == xLower & world$Y == y,]$Value)
      neighborValues = c(neighborValues, world[world$X == x & world$Y == yUpper,]$Value)
      neighborValues = c(neighborValues, world[world$X == x & world$Y == yLower,]$Value)

      
      if (includeDiagonals)
      {
        neighborValues = c(neighborValues, world[world$X == xUpper & world$Y == yUpper,]$Value)
        neighborValues = c(neighborValues, world[world$X == xUpper & world$Y == yLower,]$Value)
        neighborValues = c(neighborValues, world[world$X == xLower & world$Y == yUpper,]$Value)
        neighborValues = c(neighborValues, world[world$X == xLower & world$Y == yLower,]$Value)
      }

      currentValue = world[newWorld$X == x & newWorld$Y == y,]$Value
      newWorld[newWorld$X == x & newWorld$Y == y,]$Value = currentValue + updateFunction(currentValue, neighborValues)
    }
  }
  
  newWorld$Density = newWorld$Value / sum(newWorld$Value)
  
  return (newWorld)
}

# Update the world and display its new landscape
world = updateWorld(world, getAgentChange)
ggplot(world, aes(X, Y, z = Density)) +
  geom_raster(aes(fill = Density)) +
  geom_contour(colour = "white")
```

### Construct time data
```{r}
# Assembles the base data - what elevation is each coordinate at for each time step
buildTimeWorld <- function(width, height, iterations)
{
  lastStep = buildWorld(width, height, seq(100, 100+(height * width * 0.1), 0.1))
  lastStep$Time = rep(1, nrow(lastStep))
  
  totalData = lastStep
  
  for (t in 2:iterations)
  {
    newData = updateWorld(lastStep, getAgentChange)
    newData$Time = rep(t, nrow(newData))
    lastStep = newData
    
    totalData = rbind(totalData, newData)
  }

  return(totalData)
}

# Given a set of points over time, outputs the optimal point at each time step
getOptimalData <- function(timeWorld)
{
  optimalData = data.frame(OptimalX = 0, OptimalY = 0, Time = 0)
  i = 1
  for (t in min(timeWorld$Time):max(timeWorld$Time))
  {
    timeSubset = timeWorld[timeWorld$Time == t,]
    optimalData[i,]$OptimalX = timeSubset[timeSubset$Value == max(timeSubset$Value),][1,]$X
    optimalData[i,]$OptimalY = timeSubset[timeSubset$Value == max(timeSubset$Value),][1,]$Y
    optimalData[i,]$Time = t
    i = i + 1
  }
  return (optimalData)
}

# Given a set of points over time and an optimization function, returns the optimization function's
# guess of optimality for each time step
getGuesses <- function(timeWorld, optimizeFunction, ticksPerTime = 1, ...)
{
  guessData = data.frame(GuessedX = 0, GuessedY = 0, Time = 0)
  xGuess = floor(mean(timeWorld$X))
  yGuess = floor(mean(timeWorld$Y))
  i = 1
  for (t in min(timeWorld$Time):max(timeWorld$Time))
  {
    timeSubset = timeWorld[timeWorld$Time == t,]
    guesses = optimizeFunction(timeSubset, ticksPerTime, xGuess, yGuess, ...)
    xGuess = guesses[1]
    yGuess = guesses[2]
    guessData[i,]$GuessedX = xGuess
    guessData[i,]$GuessedY = yGuess
    guessData[i,]$Time = t
    i = i + 1
  }
  return (guessData)
}

# Horrible ugly monster function to encapsulate a bunch of work.
# Builds a world (or reuses the one passed in a timeWorld). Uses that to obtain optimal points
# (if includeOptimalPoints is true), as well as run an optimization function (if optimizeFunction is not
# null). Returns an object with three dataframes (or NULLs) - worldData, optimalPoints, and guessedPoints.
assembleData <- function (width = 17, height = 17, iterations = 50, timeWorld = NULL, optimizeFunction = NULL, ticksPerTime = 1, includeOptimalPoints = T, ...)
{
  finalData <- list(worldData = NULL, optimalPoints = NULL, guessedPoints = NULL)
  if (is.null(timeWorld))
  {
    finalData$worldData = buildTimeWorld(width, height, iterations)
  } else {
    finalData$worldData = timeWorld
  }
  if (!is.null(optimizeFunction))
  {
    finalData$guessedPoints = getGuesses(finalData$worldData, optimizeFunction, ticksPerTime, ...)
  }
  if (includeOptimalPoints)
  {
    finalData$optimalPoints = getOptimalData(finalData$worldData)
  }
  
  return (finalData)
}
```

```{r}
# Assembling a world only
finalData = assembleData(16, 16, 50)

# Save out the world for reuse
timeWorld = finalData$worldData

# See how much faster this is now?
finalData = assembleData(timeWorld = timeWorld)
```

### Prepare animated visualizer
```{r}
visualizeFunction <- function(allData, fromtime = 25, untiltime = 50)
{
  fromtime = max(min(fromtime, max(allData$worldData$Time)), min(allData$worldData$Time))
  untiltime = max(min(untiltime, max(allData$worldData$Time)), min(allData$worldData$Time))
  
  worldData = allData$worldData[allData$worldData$Time >= fromtime & allData$worldData$Time <= untiltime,]

  # Assemble base animation
  animation = ggplot(worldData, aes(X, Y, z = Density)) +
    geom_raster(aes(fill = Density)) +
    transition_states(Time, transition_length = 10, state_length = 0, wrap=F) 
    labs(title = "Time: {closest_state}") +
    enter_fade() +
    exit_fade() +
    ease_aes("linear")

  # Add optimal points, if present
  if (!is.null(allData$optimalPoints))
  {
    optData = allData$optimalPoints[allData$optimalPoints$Time >= fromtime & allData$optimalPoints$Time <= untiltime,]
    lengthenFactor = nrow(worldData) / nrow(optData)
    final = optData
    for(i in 2:lengthenFactor) final = rbind(final, optData)
    optData = final
    optData = optData[with(optData, order(Time)), ]

    animation = animation + geom_point(x=optData$OptimalX, y=optData$OptimalY, colour="hotpink", size=4)
  }
  
  # Add guessed points, if present
  if (!is.null(allData$guessedPoints))
  {
    guessData = allData$guessedPoints[allData$guessedPoints$Time >= fromtime & allData$guessedPoints$Time <= untiltime,]
    lengthenFactor = nrow(worldData) / nrow(guessData)
    final = guessData
    for(i in 2:lengthenFactor) final = rbind(final, guessData)
    guessData = final
    guessData = guessData[with(guessData, order(Time)), ]

    animation = animation + geom_point(x=guessData$GuessedX, y=guessData$GuessedY, colour="lawngreen", size=4)
  }
    
  # Display animation
  animation
}
```

### Visualize
```{r}
visualizeFunction(finalData)
```


# Setup Optimization Library
All functions have a standardized signature:

 * currentData, being the data frame for the current moment in time. Contains X, Y, and Value.
 * ticks, a measurement of how much work to do during this time interval. E.g., guesses made by simulated annealing, steps taken by hill climbing.
 * startX, the starting X value for the current run
 * startY, the starting Y value for the current run
 * additional parameters may be specified, but these must be passed into assembleData() manually
 
### Evaluation
```{r}
evaluate <- function (allData, fromtime = 25, untiltime = 50)
{
  if (is.null(allData$optimalPoints) || is.null(allData$guessedPoints))
  {
    stop("Incomplete data for evaluate; allData is missing either optimalPoints or guessedPoints")
  }
  expected = allData$optimalPoints
  actual = allData$guessedPoints
  
  fromtime = max(min(fromtime, max(expected$Time)), min(expected$Time))
  untiltime = max(min(untiltime, max(expected$Time)), min(expected$Time))
  
  xLength = max(allData$worldData$X) - min(allData$worldData$X)
  yLength = max(allData$worldData$Y) - min(allData$worldData$Y)
  
  averageDistance = 0
  for (t in fromtime:untiltime)
  {
    xLower = min(expected[expected$Time == t,]$OptimalX, actual[actual$Time == t,]$GuessedX)
    xUpper = max(expected[expected$Time == t,]$OptimalX, actual[actual$Time == t,]$GuessedX)
    yLower = min(expected[expected$Time == t,]$OptimalY, actual[actual$Time == t,]$GuessedY)
    yUpper = max(expected[expected$Time == t,]$OptimalY, actual[actual$Time == t,]$GuessedY)
    
    baseDistance = sqrt((xUpper - xLower) ^ 2 + (yUpper - yLower) ^ 2)
    xWrap = sqrt((xUpper - (xLower + xLength)) ^ 2 + (yUpper - yLower) ^ 2)
    yWrap = sqrt((xUpper - xLower) ^ 2 + (yUpper - (yLower + yLength)) ^ 2)
    bothWrap = sqrt((xUpper - (xLower + xLength)) ^ 2 + (yUpper - (yLower + yLength)) ^ 2)
    averageDistance = averageDistance + min(baseDistance, xWrap, yWrap, bothWrap)
  }
  averageDistance = averageDistance / (untiltime - fromtime + 1)
  return (averageDistance)
}
```

### Random
```{r}
randomGuesses <- function(currentData, ticks, startX = 1, startY = 1)
{
  bestX = round(runif(1, min = min(currentData$X), max = max(currentData$X)))
  bestY = round(runif(1, min = min(currentData$Y), max = max(currentData$Y)))
  return (c(bestX, bestY))
}
```


### Hill Climbing
```{r}
hillClimb <- function(currentData, ticks, startX = 1, startY = 1)
{
  bestX = startX
  bestY = startY
  bestValue = currentData[currentData$X == bestX & currentData$Y == bestY,]$Value
  for (i in 1:ticks)
  {
    xLower = bestX - 1
    if (xLower == 0) xLower = max(currentData$X)
    xUpper = bestX + 1
    if (xUpper > max(currentData$X)) xUpper = 1
    yLower = bestY - 1
    if (yLower == 0) yLower = max(currentData$Y)
    yUpper = bestY + 1
    if (yUpper > max(currentData$Y)) yUpper = 1
    
    if (currentData[currentData$X == xLower & currentData$Y == bestY,]$Value > bestValue)
    {
      bestX = xLower
      bestValue = currentData[currentData$X == xLower & currentData$Y == bestY,]$Value
    }
    
    if (currentData[currentData$X == xUpper & currentData$Y == bestY,]$Value > bestValue)
    {
      bestX = xUpper
      bestValue = currentData[currentData$X == xUpper & currentData$Y == bestY,]$Value
    }
    
    if (currentData[currentData$X == bestX & currentData$Y == yUpper,]$Value > bestValue)
    {
      bestY = yUpper
      bestValue = currentData[currentData$X == bestX & currentData$Y == yUpper,]$Value
    }
    
    if (currentData[currentData$X == bestX & currentData$Y == yLower,]$Value > bestValue)
    {
      bestY = yLower
      bestValue = currentData[currentData$X == bestX & currentData$Y == yLower,]$Value
    }
  }
  
  return (c(bestX, bestY))
}
```

### Simulated Annealing
Accepts two additional parameters:

 * coolingFunction - function that returns a percent likelihood to swap to a worse position.
 * temperatureTickAmount - each time the cooling function is called, its input increments by this amount. Determines how fast to slide along the cooling function.
```{r}
linearCoolingFunction <- function (x) { return(1 - 0.1 * x) }
powerCoolingFunction <- function (x) { return(0.75 ^ x) }
decayCoolingFunction <- function (x) { return(1 / (x + 1)) }

simulatedAnnealing <- function(currentData, ticks, startX = 1, startY = 1, coolingFunction = NULL, temperatureTickAmount = NULL)
{
  if (is.null(coolingFunction))
  {
    stop("Simulated Annealing requires CoolingFunction!")
  }
  if (is.null(temperatureTickAmount))
  {
    stop("Simulated Annealing requires TemperatureTickAmount!")
  }
  
  bestX = startX
  bestY = startY
  bestValue = currentData[currentData$X == bestX & currentData$Y == bestY,]$Value
  currentCoolingValue = 0
  for (i in 1:ticks)
  {
    xLower = bestX - 1
    if (xLower == 0) xLower = max(currentData$X)
    xUpper = bestX + 1
    if (xUpper > max(currentData$X)) xUpper = 1
    yLower = bestY - 1
    if (yLower == 0) yLower = max(currentData$Y)
    yUpper = bestY + 1
    if (yUpper > max(currentData$Y)) yUpper = 1
    
    nextX = bestX
    nextY = bestY
    nextValue = 0
    
    if (currentData[currentData$X == xLower & currentData$Y == bestY,]$Value > nextValue)
    {
      nextX = xLower
      nextValue = currentData[currentData$X == xLower & currentData$Y == bestY,]$Value
    }
    
    if (currentData[currentData$X == xUpper & currentData$Y == bestY,]$Value > nextValue)
    {
      nextX = xUpper
      nextValue = currentData[currentData$X == xUpper & currentData$Y == bestY,]$Value
    }
    
    if (currentData[currentData$X == bestX & currentData$Y == yUpper,]$Value > nextValue)
    {
      nextY = yUpper
      nextValue = currentData[currentData$X == bestX & currentData$Y == yUpper,]$Value
    }
    
    if (currentData[currentData$X == bestX & currentData$Y == yLower,]$Value > nextValue)
    {
      nextY = yLower
      nextValue = currentData[currentData$X == bestX & currentData$Y == yLower,]$Value
    }
    
    # The part that makes this simulated annealing, not hill climbing
    currentCoolingValue = currentCoolingValue + temperatureTickAmount
    if ((nextValue < bestValue && runif(1) < coolingFunction(currentCoolingValue)) ||
        nextValue > bestValue)
    {
      bestValue = nextValue
      bestX = nextX
      bestY = nextY
    }
  }
  
  return (c(bestX, bestY))
}
```

### Genetic Algorithm
```{r}
dtob <- function(number, bits = 4)
{
  return (as.integer(rev(intToBits(number)[1:bits])))
}
btod <- function(bits)
{
  return (sum(bits * 2^seq(0, length(bits) - 1, 1)))
}
getChildren <- function(parents)
{
  splitPosition = floor(runif(1, min = 0, max = length(parents[[1]])))
  child2 = c(parents[[2]][1:splitPosition], parents[[1]][(splitPosition+1):length(parents[[1]])])
  child1 = c(parents[[1]][1:splitPosition], parents[[2]][(splitPosition+1):length(parents[[2]])])
  return (list(child1, child2))
}
mutate <- function(population, mutationRate = 0.05)
{
  for (i in 1:length(population))
  {
    for (j in 1:length(population[[i]]))
    {
      if (runif(1, 0, 1) < mutationRate)
      {
        population[[i]][j] = 1 - population[[i]][j]
      }
    }
  }
  return (population)
}
getNextParents <- function(population, currentData, xBits, yBits, xMin, yMin)
{
  values = c()
  for (i in 1:length(population))
  {
    values[i] = currentData[
      currentData$X == btod(population[[i]][1:xBits])+xMin & 
      currentData$Y == btod(population[[i]][(xBits+1):(xBits+yBits)])+yMin,]$Value
  }
  return(rev(population[order(values)])[1:2])
}

geneticAlgorithm <- function(currentData, ticks, startX = 1, startY = 1, mutationRate = 0.05)
{
  xMin = min(currentData$X)
  yMin = min(currentData$Y)
  xRange = max(currentData$X) - xMin + 1
  yRange = max(currentData$Y) - yMin + 1
  xBits = ceiling(log2(xRange))
  yBits = ceiling(log2(yRange))
  
  randomParent = c(sample(c(0, 1), xBits, replace=TRUE), sample(c(0, 1), yBits, replace=TRUE))
  baseParent = c(dtob(startX - xMin, xBits), dtob(startY - yMin, yBits))
  parents = list(randomParent, baseParent)
  
  for (i in 1:ticks)
  {
    children = getChildren(parents)
    population = c(mutate(children), mutate(parents))
    parents = getNextParents(population, currentData, xBits, yBits, xMin, yMin)
  }
  
  return (c(btod(parents[[1]][1:xBits])+xMin, btod(parents[[1]][(xBits+1):(xBits+yBits)])))
}

```


### Workshop
```{r}
randomOutput = assembleData(timeWorld = timeWorld, optimizeFunction = randomGuesses, ticksPerTime = 1)
cat("Random guesses were, on average, off by", evaluate(randomOutput), "units.\n")
visualizeFunction(randomOutput)
```

```{r}
hillClimbOutput = assembleData(timeWorld = timeWorld, optimizeFunction = hillClimb, ticksPerTime = 10)
cat("Hill climbing was, on average, off by", evaluate(hillClimbOutput), "units.\n")
visualizeFunction(hillClimbOutput)
```

```{r}
saOutput = assembleData(timeWorld = timeWorld, optimizeFunction = simulatedAnnealing, ticksPerTime = 100, coolingFunction = decayCoolingFunction, temperatureTickAmount = 0.05)
cat("Simulated Annealing was, on average, off by", evaluate(saOutput), "units.\n")
visualizeFunction(saOutput)
```

```{r}
gaOutput = assembleData(timeWorld = timeWorld, optimizeFunction = geneticAlgorithm, ticksPerTime = 10, mutationRate = 0.05)
cat("Genetic Algorithm was, on average, off by", evaluate(gaOutput), "units.\n")
visualizeFunction(gaOutput)
```